---
title: "Class 10: Halloween mini project"
author: "Fan Wu(PID:A15127541)"
format: pdf
toc: true
---
## Background
As it is nearly Halloween and the half way point in the quarter, let's do a mini project to help us figure our the best candy!

## Data import

Our data comes from the 538 website and is available as a CSV file:

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
```

```{r}
head(candy)
```
```{r}
flextable::flextable(head(candy,10))
```


> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

```{r}
candy |>
  nrow()
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
My favorite winpercent

```{r}
candy["Twix", ]$winpercent
```

```{r}
## plyr is used to pull out wire by electricians

library(dplyr)
## filter() and select() are useful

candy |>
  filter(rownames(candy) == "Twix") |>
  select(winpercent)
```

```{r}
candy |>
  filter(rownames(candy) == "Almond Joy") |>
  select(winpercent)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Hershey's Special Dark", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
## Quick overview of the dataset

```{r}
## Only calling on the skim() from the package, 
## not loading other stuff from the package
skimr::skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the winpercent column

> Q7. What do you think a zero and one represent for the candy$chocolate column?

There's no chocolate in this candy

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent) +
  geom_histogram(bins = 20)
```

> Q9. Is the distribution of winpercent values symmetrical?

Not symmetrical
```{r}
ggplot(candy) + 
  aes(winpercent) +
  geom_density()
```

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

The center of distribution is below 50%, since the median is 47.83

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

On average, chocolate candy is higher ranked than fruit candy

```{r}
#1. Find all chocolate candy in the dataset

#2. Find their winpercent values

#3. Calculate the mean of these values

#4. Repeat 1,2,3 for fruity candy

#5. mean winpercent of chocolate vs. mean winpercent of fruity

choc.inds <- candy$chocolate==1
choc.win <- candy[choc.inds, ]$winpercent
choc.mean <- mean(choc.win)

choc.mean

fruit.inds <- candy$fruity==1
fruit.win <- candy[fruit.inds, ]$winpercent
fruit.mean <- mean(fruit.win)

fruit.mean
```

```{r}
# try using dplyr

candy |>
  filter(chocolate == 1) |>
  select(winpercent)

```

> Q12. Is this difference statistically significant?

Since the p-value is very small, the difference is statistically significant. 
```{r}
t.test(choc.win, fruit.win)
```

##  Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
## using dplyr
candy|>
  arrange(winpercent) |>
  head(5)
```

```{r}
# base R sort() and order()
x<- c(5, 1, 10, 4)
sort(x)
order(x)
```

```{r}
## using base R
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(-winpercent)|>
  head(5)
```


```{r}
ord.ind.top <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord.ind.top,], n=5)
```

> Q15 Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
         aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
         aes(x = winpercent, 
             y = reorder(rownames(candy), winpercent)) +
  geom_col()
```

Add some color based on the "type of candy"

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$fruity)] <- "purple"
my_cols[as.logical(candy$bar)] <- "brown"
my_cols
```

```{r}
ggplot(candy) +
         aes(x = winpercent, 
             y = reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols)
```



## Winpercent and Pricepercent

A plot with both variables/columns winpercent and pricepercent

```{r}
my_points <- rep("black", nrow(candy))
my_points[as.logical(candy$chocolate)] <- "chocolate"
my_points[as.logical(candy$fruity)] <- "purple"

```


```{r}
ggplot(candy) +
         aes(x = winpercent, 
             y = pricepercent, 
             label = rownames(candy)) +
  geom_point(col = my_points)+
  geom_text()
```

```{r}
library(ggrepel)

ggplot(candy) +
         aes(x = winpercent, 
             y = pricepercent, 
             label = rownames(candy)) +
  geom_point(col = my_points)+
  geom_text_repel(max.overlaps = 7)
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col=my_points) +
    geom_point()
```

## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
# correlation
library(corrplot)
cij <- cor(candy)
# 1 in cor() result means correlated, negative number results are negatively correlated
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

chocolate and fruity

> Q23. Similarly, what two variables are most positively correlated?

chocolate and winpercent

## Principal Component Analysis

The function to use is called `prcomp()` w/ an optional `scale=T/F` argument

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Our main PCA result figure

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols)
```

We should also examine the variable "loadings" or contributions of the original variables to the new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```

```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label = rownames(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols)
```

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and will not render to PDF :-(
```{r}
library(plotly)

# ggplotly(p)
```

